Quantum-assisted unsupervised data clustering

### 2020-04-23 manuscript.tex
Tims remarks

\documentclass[pra,showkeys,twocolumn,showpacs]{revtex4-1}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{array}
\usepackage{color}
\usepackage{soul}
% \usepackage{floatflt}
\usepackage{graphicx}
\DeclareMathOperator{\tr}{Tr}
\newcommand*{\eq}[1]{(\ref{#1})}
\newcommand*{\eqs}[2]{(\ref{#1})--(\ref{#2})}
\newcommand{\exi}{\tr{\{e^{-\frac\beta2\sum_k\lambda_k\gamma_k^\dag\gamma_k}\}}}
\renewcommand{\l}{\left(}
\renewcommand{\r}{\right)}

\begin{document}
\title{Quantum assisted unsupervised data clustering on the basis of neural networks}

@@

\begin{abstract}
	Unsupervised machine learning is one of the main directions in the area of artificial intelligence. 
	Here, we propose a realization of quantum assisted unsupervised data clustering on the basis of artificial neural network called a self-organizing feature map. 
	We make a proof-of-concept realization of one of the crucial parts of the approach on IBM quantum machine and show that it allows us to reduce number of calculations in a number of clusters. 
	We compare the results with the classical algorithm on a toy example of unsupervised text clustering.  
\end{abstract}

@@

\section{Introduction}
The combination of big data and artificial intelligence ---  dubbed the fourth industrial revolution --- has profoundly affected the modern economy in a plethora of different ways from robotics to agriculture \cite{Lecun2015, ghahramani2015,schwab2017,esteva2019, tyrsa2017}. 
Contemporary artificial intelligence methods based on neural networks also have the potential to enhance the role of novel analytical methods in science and engineering \cite{kaggle2014, radovic2018, butler2018, radovic2018}. 
Paradoxically, the exact mechanism of why neural networks are so powerful remain unknown (in many cases it is regarded as a black box).  
It has been speculated that limits to the neural network approach based on the computational power of von Neumann architecture are being approached, and improvements appear only due to heuristic escalation of complexity \cite{marcus2018,sze2017,kourtis2020}.
In particularly, the self-organizing feature map (SOFM) \cite{kohonen1990,kohonen1996,kohonen1997}, is a type of artificial neural network (ANN) that is trained in an unsupervised manner. 
SOFMs are used in many areas \cite{vilibic2016, guido1998, doszkocs1990, jones2012,mori2019,corsello2017,zhu2018,chea2016} and in comparison with many other artificial neural networks they apply competitive learning and preserve the topological properties of the input space \cite{kiviluotoa1996}. 
The SOFMs with a small number of nodes are similar to the K-means algorithm but for larger SOFMs they represent data in a fundamentally topological way that allows one to do dimensionality reduction.
Once it is trained, the map can classify a vector from the input space by finding the node with the smallest distance metric.

Meanwhile, there has been much interest recently in applying quantum computing techniques to machine learning \cite{dunjko2018, biamonte2017, schuld2014, carleo2019}. 
In the early stage, the main focus in QML was on the use of quantum computers to perform some basic linear algebra subroutines faster (such as Fourier transform and solving systems of linear equations \cite{wiebe2012,harrow2009,childs2017}), 
open up the possibilities for exponential speed-ups utilizing quantum phenomena such as quantum entanglement and quantum superposition over classical counterparts \cite{biamonte2017, schuld2014} and,
in the result, quantum versions were developed for linear regression, principal component analysis, support vector machine, K-means algorithm and so on \cite{lloyd2013,lloyd2014,dunjko2016,paparo2014,rebentrost2014}. 
Recently,  a number of shifts occurred and the hottest trend migrated to developing quantum neural networks \cite{kamruzzaman2019, schuld2014b, jeswal2019}. 
The shift was inspired with the progress in experimental quantum computing when it became possible to play around with parametrized quantum circuits, whose parameters behave much like the weights of a neural network \cite{lewenstein1994}. 
In particular, quantum algorithms for training and evaluating feedforward neural networks, one of the most usable neural network model, was developed \cite{allcock2018, tacchino2019}. 
Recent results connected classical Bayesian approach to deep learning allowed to develop a new algorithm for Bayesian deep learning on quantum computers \cite{zhao2019}. 
The quantum models for convolutional neural networks, which may be suitable for the problems of learning of quantum states were also proposed \cite{cong2019, liu2019}. Quantum classification, tested on the MNIST dataset, via Slow Feature Analysis based on the use of Quantum Frobenius Distance was proposed \cite{kerenidis2018}. 
Sublinear quantum algorithms for training linear and kernel-based classifiers were designed \cite{li2019}. 
Furthermore, continuous-variable quantum neural networks \cite{killoran2019}, quantum autoencoders \cite{bondarenko2019}, quantum approaches for reinforcement learning \cite{dunjko2017, nautrup2019, foesel2018} and some other \cite{rebentrost2018, purushothaman1997, verdon2019, cherny2019, byrnes2013, mishra2019, vinci2019, lu2019} were developed. 
Currently, it is believed that implementation of quantum neural networks can be the main test bed to achieve practical quantum supremacy on NISQ devices.

In this paper, we realize a hybrid quantum assisted SOFM (QASOFM) and apply it to the data clustering problem in an unsupervised manner on a toy example of clustering paper abstracts.   
We implement the quantum assisted SOFM in such a way that it becomes possible to use the Hamming distance as a distance metric for the training and to reduce the number of distance calculations in the number of clusters. 
We propose an optimized circuit for realizing the Hamming distance on a quantum machine. 
We prepare a toy example of data for clustering paper abstracts and give a proof-of-concept realization of the quantum assisted SOFM on the IBM Q experience quantum computer \cite{ibmq} and compare it with the classical case.

\begin{figure}
	\label{fig:sofm_fitting}
	\includegraphics[width=0.95\columnwidth]{sofm_fitting.png}
	\caption{
		Schematic illustration of the clustering problem considered in this paper.  
		Blue dots represent clusters and red dots are data points. 
		The training process moves clusters to fit the data points. 
		Note that there are fewer clusters than data points which is the essence of dimensionality reduction and is what permits the model to generalize the data.
	}
\end{figure}

@@

SOFM is one of the best-known unsupervised learning methods that is widely used in various areas of modern science. 
It was first proposed by Kohonen as a self-organizing unsupervised learning algorithm which produces feature maps similar to those occurring in the brain \cite{solan2001}. 
The SOFM algorithm operates with a set of input objects, each represented by a $N$-dimensional vector, as input and maps them onto nodes of a two-dimensional grid.

The input dimensions are associated with features and nodes in the grid, called cluster vectors, are assigned the $N$-dimensional vectors; 
the components of these vectors are usually called weights. 
Initially weight components are chosen randomly. 
We then can train our SOFM adjusting the components through the learning process which can be oversimplified into two basic procedures, 
selecting a winning cluster vector and updating its weights (Fig.~\ref{fig:sofm_fitting}). 
In more detail, they consist of four step process: 
1) select an input vector randomly from the set of all input vectors; 
2) find a cluster vector which is closest to the input vector; 
3) adjust the weights of the winning node in such a way that it becomes even closer to the input vector; 
4) repeat this process many iterations until it converges.

After the winning cluster vector is selected, the weights of the vector are adjusted according to 
%
\begin{equation}
    \label{eq:learning}
    \vec w_{\mathrm{new}} =
		\vec{w}_{\mathrm{old}} 
        + \alpha\left(\vec{x} - \vec w_{\mathrm{old}}\right)  .
\end{equation}
%
In simple words, this expression can be understood according to the following: 
if a component of the input vector is greater than the corresponding weight, increase the weight by a small amount; 
if the input component is smaller than the weight, decrease the
weight by a small amount. 
The larger the difference between the input component and the weight component, the larger increment (decrement). 
Intuitively, this procedure can be geometrically interpreted as iteratively moving the cluster vectors in space one at a time in a way that ensures each move is following the current trends inferred from their distances to the input objects. 
A visualisation of this process is provided on figure (Fig.~\ref{fig:sofm_fitting}).

@@

%%%%%
\begin{figure}[t]
	\label{fig:qcircuit}
	\includegraphics[width=0.95\columnwidth]{qcircuit.png}
	\caption{
    	%add the caption with intuitive explanation of the circuit
		A quantum circuit for the Hamming distance calculation algorithm between all samples and clusters at once. 
		First, we apply Hadamard gates in order to get superposition of the sample and cluster registers. 
		Second, we encoded information about pairwise different qubits in a quantum state of the sample register with applying the CNOT gates. 
		Third, Hamming distance values are encoded in the amplitudes of superposition with the control phase rotation and Hadamard gates. 
		Finally, a quantum state of the sample register returned to the initial basis for information retrieval.
	} 
\end{figure}
%%%%%

%%%%%
\begin{figure}[t]
	\label{fig:distance_matrix}
	\includegraphics[width=0.95\columnwidth]{distance_matrix.png}
	\caption{
		The Hamming distance matrix between two data sets of binary vectors. 
		Values of distance that less than median distance value was marked black. 
		The classical simulation of the quantum circuit shows perfect agreement to theoretical calculations and presented on the left figure. 
		Result obtained on  IBMQ "ibmq\_16\_melbourne" backend is shown on the right figure. 
		We can see good agreement between the distance matrices calculated classically and on the IBM Q Experience.
	} 
\end{figure}
%%%%%

We now introduce an optimized algorithm for calculating the matrix of Hamming distances \cite{trugenberger2001} between a sample vector and all cluster vectors at once.  
This allows for a simple encoding of the classical information into a quantum register. 
Initially we have two registers each with $n$ qubits in that we encode our set of binary vectors defined as
%
\begin{align}
    \label{eq:encodnig}
    \left| X \right\rangle  & = \sum\limits_{i=1}^{k} \left| x_i \right\rangle,  \\
    \left| Y \right\rangle&  = \sum\limits_{i=1}^{l} \left| y_i \right\rangle,
\end{align}
% 
where $ 1 \le k, l \le n $.  
Here the registers $\left| X \right\rangle$ and $\left| Y \right\rangle$ correspond to the input states and the cluster states respectively. 
The initial state of whole circuit is
%
\begin{equation} 
\left| \psi_0 \right\rangle = 
    \left| X \right\rangle
    \left| Y \right\rangle 
    \left| a \right\rangle
    \label{eq:initial_state}
\end{equation}
%
where $\left| a \right\rangle$ is an auxiliary qubit in the state $\left| 0 \right\rangle$ initially.

We then apply a CNOT gate between $\left| X \right\rangle$ and $\left| Y \right\rangle$

\begin{equation}
    \left| \psi_1 \right\rangle = 
    \mathrm{CNOT(Y,X)\left| \psi_0 \right\rangle} =  
    \sum\limits_{i, j}^{k} 
    \left| d^1_{ij}, \dots, d^n_{ij} \right\rangle 
    \left| y^1_j, \dots, y^n_j \right\rangle
    \left| 0 \right\rangle 
\end{equation}
where $d^\alpha_{ij} = \mathrm{CNOT}(y^\alpha_i, x^\alpha_j),\, \alpha = 1 \dots n$ and $i,j$ are the qubit indexes in the registers.

In the result, we encoded information about pairwise different qubits in $\{X\}$ and $\{Y\}$ into register $\left| X \right\rangle$.

After that we transfer information about distance between sample state and cluster state into the amplitude of the superposed state. 
In order to make that we apply Hadamard gate on auxiliary qubit, then control phase gate on $\left| Xa \right\rangle$,

\begin{equation}
    \label{eq:control_phase_rotation}
    R(\phi) = 
    \begin{pmatrix}
        1 & 0 & 0 & 0 \\
        0 & 1 & 0 & 0 \\
        0 & 0 & 1 & 0 \\
        0 & 0 & 0 & e^{-i\phi}
    \end{pmatrix},
    \quad \phi = \frac{\pi}{n}
\end{equation}
and again Hadamard gate.

After the first Hadamard on the ancillary qubit we have the following state

\begin{equation}
    \left| \psi_2 \right\rangle = H_a\left| \psi_1 \right\rangle = 
    \sum\limits_{i, j}^{k} 
    \left| d^1_{ij}, \dots, d^n_{ij} \right\rangle 
    \left| y^1_j, \dots, y^n_j \right\rangle
    \dfrac{\left| 0 \right\rangle + \left| 1 \right\rangle}{\sqrt{2}}
\end{equation}

Then we apply the control phase gate

\begin{multline}
    \left| \psi_3 \right\rangle = R_{(X,a)}\left(\dfrac{\pi}{n}\right)\left| \psi_2 \right\rangle
    \\ = \dfrac{1}{\sqrt{2}}
				\sum\limits_{i, j}^{k} 
				\left| d^1_{ij}, \dots, d^n_{ij} \right\rangle 
        \left| y^1_j, \dots, y^n_j \right\rangle 
        \left| 0 \right\rangle
        \\ + \dfrac{1}{\sqrt{2}}
				\sum\limits_{i, j}^{k}
        \exp\left(\dfrac{-i \pi}{n}\sum\limits_l^n d^l_{ij} \right)
        \left| d^1_{ij}, \dots, d^n_{ij} \right\rangle 
        \left| y^1_j, \dots, y^n_j \right\rangle 
        \left| 1 \right\rangle
\end{multline}

%\\ = \dfrac{1}{\sqrt{2}}
%    \sum\limits_{i, j}^{k}
%    \left| d^1_{ij}, \dots, d^n_{ij} \right\rangle 
%        \left| y^1_j, \dots, y^n_j \right\rangle 
%        \left| 0 \right\rangle
%        \\ + \dfrac{1}{\sqrt{2}}
%				\sum\limits_{i, j}^{k}
%        \left| R(d^1_{ij}, 1)\left(\dfrac{\pi}{n}\right), \dots, R(d^n_{ij}, 1)\left(\dfrac{\pi}{n}\right) \right\rangle 
%        \left| y^1_j, \dots, y^n_j \right\rangle 
%        \left| 1 \right\rangle

Applying another Hadamard on ancillary qubit and doing necessary calculations the state of quantum register reads

\begin{multline}
    \left| \psi_4 \right\rangle = 
    \sum\limits_{i, j}^{k} 
    e^{\left(\dfrac{-i \pi}{2n}\sum\limits_l^n d^l_{ij} \right)}
		\\ \times
        \left\{\cos\left(\dfrac{\pi}{2n}\sum\limits_l^n d^l_{ij} \right)
        \left| d^1_{ij}, \dots, d^n_{ij} \right\rangle 
        \left| y^1_j, \dots, y^n_j \right\rangle 
        \left| 0 \right\rangle\right.
        \\+ 
        \left. i \sin\left(\dfrac{\pi}{2n}\sum\limits_l^n d^l_{ij} \right)
        \left| d^1_{ij}, \dots, d^n_{ij} \right\rangle 
        \left| y^1_j, \dots, y^n_j \right\rangle 
        \left| 1 \right\rangle\right\}
\end{multline}

Then we return to our initial basis for register $\left| X \right\rangle$ with applying the pairwise CNOT gates:

\begin{multline}
    \left| \psi_5 \right\rangle = 
    \mathrm{CNOT(Y,X)\left| \psi_4 \right\rangle} \\=  
    \sum\limits_{i, j}^{k} 
    e^{\left(\dfrac{-i \pi}{2n}\sum\limits_l^n d^l_{ij} \right)}
				\left\{\cos\left(\dfrac{\pi}{2n}\sum\limits_l^n d^l_{ij} \right)
        \left| X_i \right\rangle 
        \left| Y_j \right\rangle 
        \left| 0 \right\rangle\right.
        \\+
        \left. i \sin\left(\dfrac{\pi}{2n}\sum\limits_l^n d^l_{ij} \right)
        \left| X_i \right\rangle 
        \left| Y_j \right\rangle 
        \left| 1 \right\rangle\right\}
\end{multline}

In the result we have the Hamming distances encoded into the amplitudes of the final state. 
After getting statistics of the measurement outcomes we can produce the distance matrix between two data sets of binary vectors (Fig.~\ref{fig:distance_matrix}). 
In this case, a biggest amplitude in measurement result coincide to a smallest Hamming distance when the measurement result of the auxiliary qubit is 0 
and we have the inverse relationship when measurement result on the auxiliary qubit is 1. 
In order to reduce noise we average measurement results over different states of the auxiliary qubit as $a_0(x_i,y_j) + (1-a_1(x_i,y_j))$ 
where $a_0(x_i,y_j)$ and $a_1(x_i,y_j)$ are amplitudes for the measurement results $\left| x_i \right\rangle\left| y_j \right\rangle \left| 0 \right\rangle_a$ and $\left| x_i \right\rangle\left| y_j \right\rangle \left| 0 \right\rangle_a$ respectively. 
In comparison with the circuits for Hamming distance calculations proposed previously \cite{trugenberger2001}, 
our method allows one to reduce number of gates in the circuit (at least in $2n$ one-qubit NOT gates, 
depends on actual realization control phase gate in a quantum register on the hardware level), 
reconstruct all distance matrix at once and can be implemented with high fidelity on current quantum devices.

>>>>

\documentclass[pra,showkeys,twocolumn,showpacs]{revtex4-1}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{array}
\usepackage{color}
\usepackage{soul}
%\usepackage{floatflt}
\usepackage{graphicx}
\DeclareMathOperator{\tr}{Tr}
\newcommand*{\eq}[1]{(\ref{#1})}
\newcommand*{\eqs}[2]{(\ref{#1})--(\ref{#2})}
\newcommand{\exi}{\tr{\{e^{-\frac\beta2\sum_k\lambda_k\gamma_k^\dag\gamma_k}\}}}
\renewcommand{\l}{\left(}
\renewcommand{\r}{\right)}

\begin{document}
\title{Quantum-assisted unsupervised data clustering}

@@

\begin{abstract}
Unsupervised machine learning is one of the main techniques in artificial intelligence. Quantum computers offer opportunities to speed up such machine learning techniques, where the number of input data elements is extremely large. Here, we propose a realization of quantum assisted unsupervised data clustering using the self-organizing feature map artificial neural networks. We make a proof-of-concept realization of one of the central components of the approach on the IBM Q Experience and show that it allows us to reduce number of calculations in a number of clusters. [TB: Explain what the main technique that is used is.  This is basically a QAOA hybrid quantum-classical approach right? If so, you should explain this] We compare the results with the classical algorithm on a toy example of unsupervised text clustering.  
\end{abstract}

@@

\section{Introduction}
The combination of big data and artificial intelligence ---  dubbed the fourth industrial revolution --- has profoundly affected the modern economy in a plethora of different ways from robotics to agriculture \cite{Lecun2015, ghahramani2015,schwab2017,esteva2019, tyrsa2017}. Contemporary artificial intelligence methods based on neural networks also have the potential to enhance the role of novel analytical methods in science and engineering \cite{kaggle2014, radovic2018, butler2018, radovic2018}. Paradoxically, the exact mechanism of why neural networks are so powerful remain unknown (in many cases it is regarded as a black box).  It has been speculated that limits to the neural network approach based on the computational power of von Neumann architecture are being approached, and improvements appear only due to heuristic escalation of complexity \cite{marcus2018,sze2017,kourtis2020}.

In particular, the self-organizing feature map (SOFM) \cite{kohonen1990,kohonen1996,kohonen1997}, is a type of artificial neural network (ANN) that is trained in an unsupervised manner. SOFMs are used in many areas \cite{vilibic2016, guido1998, doszkocs1990, jones2012,mori2019,corsello2017,zhu2018,chea2016} and in comparison with many other artificial neural networks they apply competitive learning and preserve the topological properties of the input space \cite{kiviluotoa1996}. The SOFMs with a small number of nodes are similar to the K-means algorithm but for larger SOFMs they represent data in a fundamentally topological way that allows one to do dimensionality reduction.  Once it is trained, the map can classify a vector from the input space by finding the node with the smallest distance metric.

Meanwhile, there has been much interest recently in applying quantum computing techniques to machine learning \cite{dunjko2018, biamonte2017, schuld2014, carleo2019}. The main focus of early works in quantum machine learning (QML) was on the use of quantum computers to perform basic linear algebra subroutines with a quantum speedup (such as Fourier transform and solving systems of linear equations \cite{wiebe2012,harrow2009,childs2017}) by utilizing quantum phenomena such as quantum entanglement and quantum superposition \cite{biamonte2017, schuld2014}.  Quantum versions were developed for linear regression, principal component analysis, support vector machine, K-means algorithm and others \cite{lloyd2013,lloyd2014,dunjko2016,paparo2014,rebentrost2014}. More recently, there has been attention on developing quantum neural networks \cite{kamruzzaman2019, schuld2014b, jeswal2019}. The interest in quantum neural networks was inspired by progress in experimental quantum computing when it became possible to use parametrized quantum circuits, where the parameters behave much like the weights of a neural network \cite{lewenstein1994}. In particular, quantum algorithms for training and evaluating feedforward neural networks was developed, which are one of the most usable neural network models \cite{allcock2018, tacchino2019}. Recent results connecting the classical Bayesian approach to deep learning allowed for the development of a new algorithm for Bayesian deep learning on quantum computers \cite{zhao2019}. The quantum models for convolutional neural networks, which may be suitable for the problems of learning of quantum states were also proposed \cite{cong2019, liu2019}. Quantum classification, tested on the MNIST dataset, via slow feature analysis based on the use of quantum Frobenius distance was proposed \cite{kerenidis2018}. Sublinear quantum algorithms for training linear and kernel-based classifiers were designed \cite{li2019}. Furthermore, continuous-variable quantum neural networks \cite{killoran2019}, quantum autoencoders \cite{bondarenko2019}, quantum approaches for reinforcement learning \cite{dunjko2017, nautrup2019, foesel2018} and some other \cite{rebentrost2018, purushothaman1997, verdon2019, cherny2019, byrnes2013, mishra2019, vinci2019, lu2019} were developed. Currently, it is believed that implementation of quantum neural networks can be the main test bed to achieve practical quantum supremacy on noisy intermediate scale quantum (NISQ) devices.

[TB: I think we definitely need to cite and describe hybrid quantum-classical approaches, like QAOA.  Could you please add several sentences and a solid set of references for these]

In this paper, we analyze  a hybrid quantum assisted SOFM (QASOFM) and apply it to the data clustering problem in an unsupervised manner on a toy example of clustering paper abstracts.   We implement the quantum assisted SOFM in such a way that it becomes possible to use the Hamming distance as a distance metric for the training and to reduce the number of distance calculations in the number of clusters. We propose an optimized circuit for realizing the Hamming distance on a quantum machine. We then apply it to a toy example of data for clustering paper abstracts and give a proof-of-concept realization of the quantum assisted SOFM on the IBM Q experience quantum computer \cite{ibmq} and compare it to the classical case.

\begin{figure}
  \includegraphics[width=0.95\columnwidth]{clusters.png}
\caption{Schematic illustration of the clustering problem considered in this paper.  Blue dots represent clusters and red dots are data points. The training process moves clusters to fit the data points. Note that there are fewer clusters than data points, which is the essence of dimensionality reduction, and is what permits the model to generalize the data.}
\label{clusters}
\end{figure}

@@

\subsection{The classical algorithm}

The SOFM is one of the most widely-used unsupervised learning methods used in various areas of modern science. It was first proposed by Kohonen as a self-organizing unsupervised learning algorithm which produces feature maps similar to those occurring in the brain \cite{solan2001}. The SOFM algorithm operates with a set of input objects, each represented by a $N$-dimensional vector, and maps them onto nodes of a two-dimensional grid.

The input dimensions are associated with the features, and the nodes in the grid (called cluster vectors) are assigned the $N$-dimensional vectors. The components of these vectors are usually called weights. Initially the weight components are chosen randomly. We then can train our SOFM adjusting the components through the learning process which occur in the two basic procedures of selecting a winning cluster vector and updating its weights (Fig.~\ref{clusters}). More specifically, they consist of four step process: 1) selecting an input vector randomly from the set of all input vectors; 2) finding a cluster vector which is closest to the input vector; 3) adjusting the weights of the winning node in such a way that it becomes even closer to the input vector; 4) repeating this process for many iterations until it converges.

After the winning cluster vector is selected, the weights of the vector are adjusted according to [TB: all variables should be defined: what is $ \vec{x} , \alpha $ etc. ] [TB: this is done iteratively? Then should there be an index over  $ \vec{x}_i $? ]
%
\begin{equation}
    \vec w_{\mathrm{new}} =
        \vec w_{\mathrm{old}} 
        + \alpha\left(\vec{x} - \vec w_{\mathrm{old}}\right)  .
    \label{eq:learning}
\end{equation}
%
The above expression can be interpreted according to: if a component of the input vector is greater than the corresponding weight, increase the weight by a small amount; if the input component is smaller than the weight, decrease the
weight by a small amount. The larger the difference between the input component and the weight component, the larger the increment (decrement). Intuitively, this procedure can be geometrically interpreted as iteratively moving the cluster vectors in space one at a time in a way that ensures each move is following the current trends inferred from their distances to the input objects. A visualisation of this process is shown in Fig. \ref{clusters}.

@@

%%%%%
\begin{figure}[t]
\includegraphics[width=0.95\columnwidth]{circuit.png}
\caption{
    Quantum circuit for the quantum parallelized Hamming distance calculation between all samples and clusters. First,  Hadamard gates are applied in order to obtain a superposition of the sample and cluster registers. Second, we encoded information about pairwise different qubits in a quantum state of the sample register with applying the CNOT gates. Third, Hamming distance values are encoded in the amplitudes of superposition with the control phase rotation and Hadamard gates. Finally, a quantum state of the sample register returned to the initial basis for information retrieval. [TB: The figure is too small to be read, please make this bigger]
} 
\label{circuit}
\end{figure}
%%%%%

%%%%%
\begin{figure}[t]
\includegraphics[width=0.95\columnwidth]{distance_matrix.png}
\caption{Hamming distance matrix between two data sets of binary vectors. Values of distance that are less than the median distance value are marked black. Classical simulation of the quantum circuit shows perfect agreement to theoretical calculations and presented on the left figure. Result obtained on the IBM Q Experience ``ibmq\_16\_melbourne'' backend is shown on the right figure. We can see good agreement between the distance matrices calculated classically and on the IBM Q Experience.  [TB: could you put (a) (b) labels as with the other figures. ]} 
\label{fig:distance_matrix}
\end{figure}
%%%%%

We now introduce an optimized algorithm for calculating the matrix of Hamming distances \cite{trugenberger2001} between a sample vector and all cluster vectors, making use of quantum parallelism.  This allows for a simple encoding of the classical information into a quantum register. [TB: allows?]

The overall procedure involves two registers of $n$ qubits each, denoted $\left| X \right\rangle$ and $\left| Y \right\rangle$, along with a single auxiliary qubit $\left| a \right\rangle$. During the whole process, the $\left| Y \right\rangle$ register is used to store the cluster states.  At the beginning and end of the procedure the $\left| X \right\rangle$ register stores the input vectors.  Meanwhile, during the procedure it stores the differences between input vectors and cluster states.

Let us assume we have $k$ input vectors and $l$ cluster states. The $i$th input vector and $j$th cluster vector are respectively denoted as $\left| x_i \right\rangle$, $\left| y_j \right\rangle$. The registers $\left| X \right\rangle$ and $\left| Y \right\rangle$ are initialized to store the input vectors and cluster vectors according to
%
\begin{align}
    \left| X \right\rangle  & = \frac{1}{\sqrt{k}} \sum\limits_{i=1}^{k} \left| x_i \right\rangle,  \\
    \left| Y \right\rangle&  = \frac{1}{\sqrt{l}} \sum\limits_{j=1}^{l} \left| y_j \right\rangle .
    \label{eq:encodnig}
\end{align}
% 
The two registers along with the auxiliary qubit comprise the initial state of the quantum computer according to
%
\begin{equation} 
\left| \psi_0 \right\rangle = 
    \left| X \right\rangle
    \left| Y \right\rangle 
    \left| a \right\rangle
    \label{eq:initial_state}
\end{equation}
%
where $\left| a \right\rangle$ is an auxiliary qubit in the state $\left| 0 \right\rangle$ initially.

Given this initial state we may begin the processing of the problem. We start by applying a CNOT gate between $\left| X \right\rangle$ and $\left| Y \right\rangle$

\begin{align}
    | \psi_1 \rangle & = 
    \mathrm{CNOT} (Y,X)| \psi_0 \rangle \nonumber \\
		& =  
    \frac{1}{\sqrt{kl}} \sum_{i, j=1}^{k} 
    | d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \rangle 
    | y^{(1)}_j, \dots, y^{(n)}_j \rangle
    | 0 \rangle 
\end{align}
%
where $d^\alpha_{ij} = \mathrm{CNOT}(y^\alpha_i, x^\alpha_j)$, $\alpha = 1 \dots n$, and $i,j$ are the qubit indexes in the registers. At this stage of the computation the $\left| X \right\rangle$ no longer stores the input vectors, instead it stores the information about pairwise different qubits between the input vector $\{X\}$ and cluster vector $\{Y\}$. Next, for each pair $\{X\}$ and $\{Y\}$, the accumulated information of all the differences is projected onto the amplitude of the superposed state. This is achieved by applying the Hadamard gate on auxiliary qubit, followed by a controlled phase gate on $\left| Xa \right\rangle$ defined as
%
\begin{equation}
    \label{eq:control_phase_rotation}
    R(\phi) = 
    \begin{pmatrix}
        1 & 0 & 0 & 0 \\
        0 & 1 & 0 & 0 \\
        0 & 0 & 1 & 0 \\
        0 & 0 & 0 & e^{-i\phi}
    \end{pmatrix} .
    \quad \phi = \frac{\pi}{n}
\end{equation}
%
Finally, another Hadamard gate is applied. [TB: on which qubit? I think its better just to describe this using a circuit, rather than describe the steps in words. ]

After the first Hadamard on the ancilla qubit the state is
%
\begin{equation}
    \left| \psi_2 \right\rangle = H_a\left| \psi_1 \right\rangle = 
    \sum\limits_{i, j=1}^{k} 
    \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
    \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle
    \dfrac{(\left| 0 \right\rangle + \left| 1 \right\rangle)}{\sqrt{2}}  .
\end{equation}
%
Applying the controlled phase gate the state then becomes [TB: please define all variables like $ R_{(X,a)} $]
%
\begin{multline}
    \left| \psi_3 \right\rangle = R_{(X,a)}\left(\dfrac{\pi}{n}\right)\left| \psi_2 \right\rangle
    \\ = \dfrac{1}{\sqrt{2}}
				\sum\limits_{i, j=1}^{k} 
				\left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 0 \right\rangle
        \\ + \dfrac{1}{\sqrt{2}}
				\sum\limits_{i, j=1}^{k}
        \exp\left(\dfrac{-i \pi}{n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
\\ \times        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 1 \right\rangle
\end{multline}
%
%
%\\ = \dfrac{1}{\sqrt{2}}
%    \sum\limits_{i, j}^{k}
%    \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
%        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
%        \left| 0 \right\rangle
%        \\ + \dfrac{1}{\sqrt{2}}
%				\sum\limits_{i, j}^{k}
%        \left| R(d^{(1)}_{ij}, 1)\left(\dfrac{\pi}{n}\right), \dots, R(d^{(n)}_{ij}, 1)\left(\dfrac{\pi}{n}\right) \right\rangle 
%        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
%        \left| 1 \right\rangle 
%
Applying another Hadamard on the ancilla qubit we obtain
%
\begin{multline}
    \left| \psi_4 \right\rangle = 
    \sum\limits_{i, j=1}^{k} 
    \exp \left(\dfrac{-i \pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
		\\ \times
        \left[ \cos\left(\dfrac{\pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 0 \right\rangle\right.
        \\+ 
        \left. i \sin\left(\dfrac{\pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 1 \right\rangle\right] .
\end{multline}
%
This completes the step for projecting differences between pairs of $\{X\}$ and $\{Y\}$ onto the amplitude of the auxiliary qubit. The process is done in the $x$-basis, achieved by the surrounding Hadamard gates. There are two possible measurement outcomes of the auxiliary qubit.  Each pair of $\{X\}$ and $\{Y\}$ forms a subspace of the Hilbert space, the controlled phase gate ensures to change amplitudes of those outcomes within this subspace depending on how different the spin configurations between $\{X\}$ and $\{Y\}$ are.

At this stage, the information regarding the differences between pairs of $\{X\}$ and $\{Y\}$ is no longer relevant, thus we return to our initial basis for register $\left| X \right\rangle$ by applying pairwise CNOT gates:
%
\begin{multline}
    \left| \psi_f \right\rangle = 
    \mathrm{CNOT} (Y,X)\left| \psi_4 \right\rangle \\=  
    \sum\limits_{i, j=1}^{k} 
    \exp \left(\dfrac{-i \pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
				\left[ \cos\left(\dfrac{\pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| X_i \right\rangle 
        \left| Y_j \right\rangle 
        \left| 0 \right\rangle\right.
        \\+
        \left. i \sin\left(\dfrac{\pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| X_i \right\rangle 
        \left| Y_j \right\rangle 
        \left| 1 \right\rangle\right] .
\end{multline}
%
This makes $\{X\}$ store the input vectors again, as in the initial step. It however preserves the amplitudes of the auxiliary qubit which are proportional to how different each pairs of $\{X\}$ and $\{Y\}$ are.

We thus have the Hamming distances encoded into the amplitudes of the final state. From the statistics of the measurement outcomes, the distance matrix between two data sets of binary vectors can be obtained (Fig.~\ref{fig:distance_matrix}). In this case, the biggest amplitude of the measurement result coincides with the smallest Hamming distance when the measurement result of the ancilla qubit is 0. 
If the ancilla qubit is 1, there is an inverse relationship with the measurement result.  [TB: what kind of inverse relationship?]
